Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-05 analysis/paper/paper.Rmd
updates for intro

The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society. Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations. According to the political approaches, craft specialization, attached production specifically, is viewed as a strategy used by elites to control over production systems for the sake of power and authority [@Costin2001]. The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economic]. However, the explanation for the emergence of specialized production might not be exclusive or straightforward. Furthermore, specialization is not necessarily accompanied with social differentiation observed in lowland northern Philippines with a low degree of ceramic specialization in Ifugao societies with social ranking during the Spanish colonization [@Acabado2018]. This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2015; @Costin2001].

@@

To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003; @Roux2018; @Stark1995]. However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assemblages. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Using geometric morphomerics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.

@@

## social changes in cross-cultural interaction contexts
- cultural resistance
- trade [@Alizadeh2018]
- identity, social boundary, Raux2015, Boness2017

>>>>

The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society. Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations. According to the political approaches, craft specialization, attached production specifically, is viewed as a strategy used by elites to control over production systems for the sake of power and authority [@Costin2001]. The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. Economic models consider production, exchange, and consumption as a whole to explain specialization as a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], a reaction to the changes in subsistence practices [@Stark1995economic], or an outcome stimulated by long-distance trade and inter-regional exchange [@Alizadeh2018]. An present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating the correlation with major socio-economic changes [@Roux2015]. However, it is important to note that the explanation for the emergence of specialized production might not be exclusive or straightforward. Moreover, @Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on the result of a low degree of ceramic specialization in Ifugao ranked societies during the Spanish colonization. The development of specialization could be history dependent, where condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2015; @Costin2001].

@@

To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison to identify lower value as an indicator of higher degree of standardization [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region. Although this traditional typological and linear measurements are useful for identifying differences in archaeological assembles, it might be limited because dimensional measurements can be insensitive to subtle variations resulting from changes in assemblages. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Using geometric morphomerics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.

@@

Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape throughout 500 years from a single site that can represent the development of social organization in northeastern Taiwan? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?

- cultural resistance
- trade [@Alizadeh2018]
- identity, social boundary, Raux2015, Boness2017
